Implementing Distributed Systems Directly with I/O Automata in Ioa++

### 2011-09-23 papers/paper1/design.tex
Removed listings and revised evaluation for 1000 trials.

An implementation of the component model described in Section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
The Scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
The services provided by the base class are indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}.
The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a User Automaton.
An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.

@@

\subsection{Implementation}

Ioa++ is implemented in C++ for POSIX environments and is available at \url{http://jrwilson.github.com/ioa/}.

>>>>

Ioa++ is an implementation in C++ of the component model described in Section~\ref{component_model} for POSIX environments and is available at \url{http://jrwilson.github.com/ioa/}.
Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.
The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
The Scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
The services provided by the base class are indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}.
The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a User Automaton.
An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.

@@

### 2011-09-23 papers/paper1/evaluation.tex
Removed listings and revised evaluation for 1000 trials.

A trial consists of a choice for $\rho$, $N$, and the number of actions to be executed.
For the experiments, the parameter $\rho$ was varied from $0.0$ to $1.0$ in increments of $0.1$ and the parameter $N$ was varied from $1$ to $1000000$ by factors of $10$.
The number of actions executed by each automaton was fixed at 1000.
Each trial was repeated 30 times.
All calculations were performed assuming a normal distribution at a confidence level of 95\%.

The driver for a single trial can be found in examples/random.cpp of the ioa++ package.
The trials were performed on a MacBook Pro running OS X 10.5.8 with an 2.4 GHz Intel Core 2 Duo processor with 4GB of memory.
The code was compiled and linked with i686-apple-darwin9-g++-4.0.1 build 5493 with -O2 optimization and the -DPROFILE flag to enable profiling in ioa++.

@@

Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with one and two threads.
When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
Consequently, every action must be serialized yielding a maximum speed-up of 1.
When $\rho = 1$, every action is an internal action and can be executed concurrently with a corresponding maximum speed-up of 2.

@@

Figures~\ref{overhead1} and \ref{overhead2} show the per thread average of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions.
(Note that the confidence intervals are present but small.)
Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
The fraction decreases as $\rho$ goes to 1 as the system transitions from acquiring two locks to one lock.
The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends virtually no time on synchronization when $\rho = 1$.

This experiment indicates that concurrent execution is possible with ioa++ but a speed-up will only be achieved if (1) enough independent actions are enabled and (2) the duration of the independent actions overcome the overhead of ioa++.
The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the ioa::schedule call.
From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4211ns $\pm$ 31ns for the single threaded experiments and 4858ns $\pm$ 32ns for the multi-threaded experiments.
Similary, we calculated the scheduling overhead and found it to be 227ns $\pm$ 2.5ns for the single threaded experiments and 279ns $\pm$ 2.3ns for the multi-threaded experiments.
Both of these can be improved with better scheduler design.
The synchronization time depends on the interactions among the system being executed, the scheduler, and the locking scheme used by the framework.
We plan to address scheduling and synchronization in future work.

>>>>

A trial consists of a choice for $\rho$, $N$, and the number of actions to be executed.
For the experiments, the parameter $\rho$ was varied from $0.0$ to $1.0$ in increments of $0.1$ and the parameter $N$ was varied from $1$ to $1000000$ by factors of $10$.
The number of actions executed by each automaton was fixed at 1000.
Each trial was repeated 1000 times.
All calculations were performed assuming a normal distribution at a confidence level of 95\%.

The driver for a single trial can be found in examples/random.cpp of the ioa++ package.
The trials were performed on a Mac G4 running OS X 10.5.8 with two 2.66 GHz Dual-Core Intel Xeon processors with 2GB of memory.
The code was compiled and linked with i686-apple-darwin9-g++-4.0.1 build 5493 with -O2 optimization and the -DPROFILE flag to enable profiling in ioa++.

@@

Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with one and two threads.
Confidence intervals are shown with largest being 0.0165.
When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
Consequently, every action must be serialized yielding a maximum speed-up of 1.
When $\rho = 1$, every action is an internal action and can be executed concurrently with a corresponding maximum speed-up of 2.

@@

Figures~\ref{overhead1} and \ref{overhead2} show the per thread average of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions.
Confidence intervals are shown with largest being 0.002155 for Figure~\ref{overhead1} and 0.001781 for Figure~\ref{overhead2}.
Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
The fraction decreases as $\rho$ goes to 1 as the system transitions from acquiring two locks to one lock.
The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends virtually no time on synchronization when $\rho = 1$.

This experiment indicates that concurrent execution is possible with ioa++ but a speed-up will only be achieved if (1) enough independent actions are enabled and (2) the duration of the independent actions overcome the overhead of ioa++.
The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the ioa::schedule call.
From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4316ns $\pm$ 3.5ns for the single threaded experiments and 6067ns $\pm$ 3.6ns for the multi-threaded experiments.
Similary, we calculated the scheduling overhead and found it to be 247ns $\pm$ 0.678ns for the single threaded experiments and 307ns $\pm$ 0.357ns for the multi-threaded experiments.
Both of these can be improved with better scheduler design.
The synchronization time depends on the interactions among the system being executed, the scheduler, and the locking scheme used by the framework.
We plan to address scheduling and synchronization in future work.

### 2011-09-23 papers/paper1/programming_model.tex
Removed listings and revised evaluation for 1000 trials.

\paragraph*{Declaring an automaton}
All automata inherit from the common base class ioa::automaton which implements the system action interface.
The following code declares the asynch\_lcr\_automaton:
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
The asynch\_lcr\_automaton is declared as a template where the UID template parameter must model the unique identifier concept.
The generic programming facilities of the C++ language aid in the development of general purpose reusable components.

\paragraph*{Declaring and initializing state variables}
The state variables of an automaton are expressed as member variables of the corresponding class.
The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status:
\begin{lstlisting}
private:
  const UID m_u;
  std::queue<UID> m_send;
  bool m_report;
  bool m_leader;
\end{lstlisting}
Notice that the member variables are declared private which is in keeping with the notion that the state of each automaton is independent.
State variables are initialized using a constructor.
For the asynch\_lcr\_automaton, the flags are both false and the UID is initialized and added to the send queue:
\begin{lstlisting}
public:
  asynch_lcr_automaton (const UID& u) :
    m_u (u),
    m_report (false),
    m_leader (false)
  {
    m_send.push (m_u);
    schedule ();
  }
\end{lstlisting}
The constructor also bootstraps the scheduler by scheduling all enabled local actions using the member function schedule.
The schedule member function is discussed in the paragraph on scheduling.

@@

To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
Local actions are declared using three functions corresponding to a precondition, effect, and scheduling call.
The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
For example, the send action of the asynch\_lcr\_automaton is declared:
\begin{lstlisting}
private:
  bool send_precondition () const { ... }
  UID send_effect () { ... }
  void send_schedule () const { ... }
public:
  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
\end{lstlisting}
Input actions are declared similarly save the absence of a precondition.
The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
The macros for actions that are subject to binding by other automata belong in a public section.

\paragraph*{Scheduling}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
For the asynch\_lcr\_automaton, the scheduling function is:
\begin{lstlisting}
  void schedule () const {
    if (send_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::send);
    }
    if (leader_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::leader);
    }
  }
\end{lstlisting}
The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this function.

@@

\paragraph*{Dynamic composition}
The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
Children automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a generator for allocating the child automaton.
Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class contain methods for destroying and unbinding respectively.
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.

\paragraph*{Binding predicates}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of actions bound to the specified action.
The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, ioa::binding\_count is used in the precondition of outputs to ensure that the value generated by the output will be received.
This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
\begin{lstlisting}
bool send_precondition () const {
  return !m_send.empty () &&
    ioa::binding_count (&asynch_lcr_automaton::send)
      != 0;
}
\end{lstlisting}

>>>>

\paragraph*{Declaring an automaton}
All automata inherit from the common base class ioa::automaton which implements the system action interface.
The following code declares the asynch\_lcr\_automaton:
\ifjournal
\begin{lstlisting}
template <typename UID>
class asynch_lcr_automaton :
  public ioa::automaton
{
  ...
};
\end{lstlisting}
\fi
The asynch\_lcr\_automaton is declared as a template where the UID template parameter must model the unique identifier concept.
The generic programming facilities of the C++ language aid in the development of general purpose reusable components.

\paragraph*{Declaring and initializing state variables}
The state variables of an automaton are expressed as member variables of the corresponding class.
The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status:
\ifjournal
\begin{lstlisting}
private:
  const UID m_u;
  std::queue<UID> m_send;
  bool m_report;
  bool m_leader;
\end{lstlisting}
\fi
Notice that the member variables are declared private which is in keeping with the notion that the state of each automaton is independent.
State variables are initialized using a constructor.
For the asynch\_lcr\_automaton, the flags are both false and the UID is initialized and added to the send queue:
\ifjournal
\begin{lstlisting}
public:
  asynch_lcr_automaton (const UID& u) :
    m_u (u),
    m_report (false),
    m_leader (false)
  {
    m_send.push (m_u);
    schedule ();
  }
\end{lstlisting}
\fi
The constructor also bootstraps the scheduler by scheduling all enabled local actions using the member function schedule.
The schedule member function is discussed in the paragraph on scheduling.

@@

To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
Local actions are declared using three functions corresponding to a precondition, effect, and scheduling call.
The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
For example, the send action of the asynch\_lcr\_automaton is declared:
\ifjournal
\begin{lstlisting}
private:
  bool send_precondition () const { ... }
  UID send_effect () { ... }
  void send_schedule () const { ... }
public:
  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
\end{lstlisting}
\fi
Input actions are declared similarly save the absence of a precondition.
The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
The macros for actions that are subject to binding by other automata belong in a public section.

\paragraph*{Scheduling}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
For the asynch\_lcr\_automaton, the scheduling function is:
\ifjournal
\begin{lstlisting}
  void schedule () const {
    if (send_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::send);
    }
    if (leader_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::leader);
    }
  }
\end{lstlisting}
\fi
The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this function.

@@

\paragraph*{Dynamic composition}
The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\ifjournal
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
Children automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a generator for allocating the child automaton.
Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class contain methods for destroying and unbinding respectively.
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.

\paragraph*{Binding predicates}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of actions bound to the specified action.
The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, ioa::binding\_count is used in the precondition of outputs to ensure that the value generated by the output will be received.
This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
\ifjournal
\begin{lstlisting}
bool send_precondition () const {
  return !m_send.empty () &&
    ioa::binding_count (&asynch_lcr_automaton::send)
      != 0;
}
\end{lstlisting}
\fi
